Reduced Nucleotide Diversity on a Plant Y Chromosome Following Recent Recombination Suppression

### 2016-09-21 manuscript.tex
stephen's changes + minor edits

In addition to demographic factors, evolutionary theory predicts that genetic variability on sex chromosomes should be affected by the evolution of suppressed X-Y recombination, which lowers the coalescent $N_{e}$ for Y-linked genes and increases their vulnerability to the effects of selective sweeps \citep{smith1974hitch,aquadro1994} and background selection \citep{charlesworth1996background,charlesworth1994effect}. These processes can be viewed as examples of the Hill–Robertson (HR) effect, in which the effective population size for a given chromosomal region depends strongly on the rate of recombination, such that selection at a focal site is not independent of selection at nearby sites \citep{hill1966HReffect}. Moreover, both background selection and selective sweeps predict a reduction in $N_{e}$ and, as a result, reduced levels of neutral polymorphism, selection efficacy, and rate of adaptation \citep{comeron2008}. Although the HR effect can occur throughout the genome \citep{mcvean2000}, large genomic regions in which recombination is suppressed, such as the Y chromosome, are expected to experience the most severe reductions in neutral diversity due to chromosome-wide linkage \citep{charlesworth1996CB,charlesworth2000degeneration,bachtrog2013NRG}. Consistent with this, studies examining Y chromosome variability in mammals \citep{hellborg2004,Wilsonsayres2014} and Drosophila \citep{mcallister1999,bachtrog2000} have found that Y-linked diversity is considerably lower than predicted under models of neutral evolution, suggesting that the HR effect plays a key role determining patterns of nucleotide diversity during sex chromosome evolution across a broad range of animal taxa.

Despite widespread interest in determining the effects of recombination on patterns of genetic diversity \citep{ellegren2011,bachtrog2013NRG}, we still know very little about the influence of the HR effect on patterns nucleotide polymorphism in younger sex chromosome systems, where \textit{de novo} recombination suppression evolved relatively recently \citep{charlesworth2016plant}. Recent studies of young plant sex chromosome systems have revealed that several canonical features of X and Y chromosome evolution are shared between plants and animals, including the genetic deterioration of the Y chromosome \citep{bergero2015,hough2014}, the occurrence of ‘evolutionary strata’ \citep{bergero2009}, and the evolution of dosage compensation \citep{muyle2012,papadopulos2015}. Nonetheless, except for a small number of genes in \textit{Silene latifolia} \citep{filatov2001diversity,qiu2010nucleotide}, the associated changes in genetic diversity that are predicted from evolutionary models of X- and Y-chromosome evolution have not been widely studied in plant sex chromosomes. Given that background selection and selective sweeps are predicted to have the strongest effects during the earliest stages of Y-chromosome evolution, before Y-chromosomes have lost the majority of their genes \citep{bachtrog2008temporal}, studying recently evolved sex chromosomes provides an important test of the temporal dynamics of the HR effect.

@@

Finally, we tested for population substructure between the two sex chromosome races to control for the possibility that hidden substructure in our sampled populations could affect estimates of diversity. To do this, we conducted a phylogenetic analysis of sex-linked sequences from all populations in our study. The analysis was conducted on an alignment of the X- and Y-linked genes from \textit{R. hastatulus}, with orthologous autosomal sequences from the non-dioecious but closely related outgroup species \textit{R. bucephalophorus} used to root the tree (Figure S2). The phylogeny was inferred using the Neighbor-Joining method \citep{saitou1987neighbor}, with evolutionary distances computed using Maximum Composite Likelihood \citep{tamura2011mega5}. The inferred sequence phylogeny revealed strong support for Y-linked genes from the $XY_{1}XY_{2}$ ("North Carolina") race being paraphyletic (98\% bootstrap support), with samples from two populations (hereafter the "SC sub-clade") forming a monophyletic group, and samples from Florida and Georgia (hereafter the "FL sub-clade") also forming a monophyletic group, that was more closely related to the XY ("Texas") race (Figure S2). In light of this evidence for significant population substructure, and to consider between population differences, we estimated X and Y diversity in each of the three sub-clades (SC, FL, TX) separately.

@@

For purifying selection simulations, we assumed that selection coefficients followed a gamma distribution, and we considered a range of values for the mean selection coefficient against deleterious mutations, $s$, by varying the scale parameter from 0.0001 to 0.1 while keeping the distribution shape parameter constant (see Supporting Information for simulation commands). To make the simulation output comparable to our data, we initialized the simulations with our empirically estimated autosomal $\theta$, scaled to the Y chromosome as $\theta_{Y}=\theta_{A}/4$ and adjusted for a sex ratio of $N_{m}/N_{f}=0.4$, and we sampled 8 haploid chromosomes per simulation. Simulated sequences contained 500 kb of linked neutral sequence from which we calculated diversity.

>>>>

In addition to demographic factors, evolutionary theory predicts that genetic variability on sex chromosomes should be affected by the evolution of suppressed X-Y recombination, which lowers the coalescent $N_{e}$ for Y-linked genes and increases their vulnerability to the effects of selective sweeps \citep{smith1974hitch,aquadro1994} and background selection \citep{charlesworth1996background,charlesworth1994effect}. Moreover, both background selection and selective sweeps predict a reduction in $N_{e}$ and, as a result, reduced levels of neutral polymorphism, selection efficacy, and rate of adaptation \citep{comeron2008}. Although the such linked selective effects can occur throughout the genome \citep{mcvean2000}, large genomic regions in which recombination is suppressed, such as the Y chromosome, are expected to experience the most severe reductions in neutral diversity due to chromosome-wide linkage \citep{charlesworth1996CB,charlesworth2000degeneration,bachtrog2013NRG}. Consistent with this, studies examining Y chromosome variability in mammals \citep{hellborg2004,Wilsonsayres2014} and Drosophila \citep{mcallister1999,bachtrog2000} have found that Y-linked diversity is considerably lower than predicted under models of neutral evolution, suggesting that the linked selection plays a key role determining patterns of nucleotide diversity during sex chromosome evolution across a broad range of animal taxa.

Despite widespread interest in determining the effects of recombination on patterns of genetic diversity \citep{ellegren2011,bachtrog2013NRG}, we still know very little about the influence of linked selection on patterns nucleotide polymorphism in younger sex chromosome systems, where \textit{de novo} recombination suppression evolved relatively recently \citep{charlesworth2016plant}. Recent studies of young plant sex chromosome systems have revealed that several canonical features of X and Y chromosome evolution are shared between plants and animals, including the genetic deterioration of the Y chromosome \citep{bergero2015,hough2014}, the occurrence of ‘evolutionary strata’ \citep{bergero2009}, and the evolution of dosage compensation \citep{muyle2012,papadopulos2015}. Nonetheless, except for a small number of genes in \textit{Silene latifolia} \citep{filatov2001diversity,qiu2010nucleotide}, the associated changes in genetic diversity that are predicted from evolutionary models of X- and Y-chromosome evolution have not been widely studied in plant sex chromosomes. Given that background selection and selective sweeps are predicted to have the strongest effects during the earliest stages of Y-chromosome evolution, before Y-chromosomes have lost the majority of their genes \citep{bachtrog2008temporal}, studying recently evolved sex chromosomes provides an important test of the temporal dynamics of the linked selection.

@@

Finally, we tested for population substructure within and between the two sex chromosome races to control for the possibility that hidden substructure in our sampled populations could affect estimates of diversity. To do this, we conducted a phylogenetic analysis of sex-linked sequences from all populations in our study. The analysis was conducted on an alignment of the X- and Y-linked genes from \textit{R. hastatulus}, with orthologous autosomal sequences from the non-dioecious but closely related outgroup species \textit{R. bucephalophorus} used to root the tree (Figure S2). The phylogeny was inferred using the Neighbor-Joining method \citep{saitou1987neighbor}, with evolutionary distances computed using Maximum Composite Likelihood \citep{tamura2011mega5}. The inferred sequence phylogeny revealed strong support for Y-linked genes from the $XY_{1}XY_{2}$ ("North Carolina") race being paraphyletic (98\% bootstrap support), with samples from two populations (hereafter the "SC sub-clade") forming a monophyletic group, and samples from Florida and Georgia (hereafter the "FL sub-clade") also forming a monophyletic group, that was more closely related to the XY ("Texas") race (Figure S2). In light of this evidence for significant population substructure, and to consider between population differences, we estimated X and Y diversity in each of the three sub-clades (SC, FL, TX) separately.

@@

For purifying selection simulations, we assumed that selection coefficients followed a gamma distribution, and we considered a range of values for the mean selection coefficient against deleterious mutations, $s$, by varying the scale parameter from 0.0001 to 0.1 while keeping the distribution shape parameter constant (see Supporting Information for simulation commands). To make the simulation output comparable to our data, we initialized the simulations with our empirically estimated autosomal $\theta$, adjusted for a sex ratio of $N_{m}/N_{f}=0.4$, and we sampled 8 haploid chromosomes per simulation. Simulated sequences contained 500 kb of linked neutral sequence from which we calculated diversity.
